Component response rate variation underlies the stability of complex systems

### 2020-03-11 notebook/ms.tex
Smooth the text to compare matrices rather than variances in component response rate (which was a bit more difficult to read

\textbf{Numerical simulations of random \(\mathbf{M}\) with and without
\(\mathbf{\sigma^{2}_{\gamma}}\)}. I used numerical simulations and
eigenanalysis to test how variation in \(\gamma\) affects stability in
random matrices with known properties, comparing the stability of
\(\textbf{A}\) versus \(\mathbf{M} = \gamma\mathbf{A}\). Values of
\(\gamma\) were sampled from a uniform distribution where
\(\gamma_{i} \sim \mathcal{U}(0, 2)\) and \(\sigma^{2}_{\gamma} = 1/3\)
(see Supplementary Information for other distributions, which give
similar results). Diagonal elements of simulated \(\mathbf{A}\) were
standardised to the mean value of sampled \(\gamma\) so that \(E[d]\) of
\(\textbf{M}\) and \(\textbf{M}\) were guaranteed to be identical. Here
I focus on the effect of \(\gamma\) across values of \(\rho\), and for
increasing system sizes (\(S\)) in random and structured networks. By
increasing \(S\), the objective is to determine the effect of \(\gamma\)
as system complexity increases toward the boundary at which stability is
realistic for a finite system.

@@

Ten thousand numerical simulations of \(\mathbf{M}\) revealed that
\(\sigma^{2}_{\gamma} > 0\) results in a nonlinear relationship between
\(\rho\) and \(\Re(\lambda_{max})\), which can sometimes increase the
stability of the system. Figure 2 shows a comparison of
\(\Re(\lambda_{max})\) across \(\rho\) values when
\(\sigma^{2}_{\gamma} = 0\) versus \(\sigma^{2}_{\gamma} = 1/3\) given
\(S = 25\), \(C = 1\), and \(\sigma = 0.4\). For
\(-0.4 \leq \rho \leq 0.7\) (shaded region of Fig. 2), expected
\(\Re(\lambda_{max})\) was lower given \(\sigma^{2}_{\gamma} = 1/3\).
For \(\rho \geq -0.1\), the range of \(\Re(\lambda_{max})\) values
observed given \(\sigma^{2}_{\gamma} = 1/3\) also had a lower bound,
which included negative values for \(\rho = -0.1\) and \(\rho = 0\).
Hence, across a wide range of system correlations, variation in the
response rate of system components had a stabilising effect.

The stabilising effect of \(\sigma^{2}_{\gamma}\) across \(\rho\)
increased with increasing \(S\). Figure 3 shows 10000 numerical
simulations of \(\mathbf{M}\) for \(\sigma^{2}_{\gamma} = 0\) versus
\(\sigma^{2}_{\gamma} = 1/3\) across increasing \(S\) given \(C = 1\)
and \(\sigma = 0.2\) (\(\sigma\) has been lowered here to better
illustrate the effect of \(S\); note that now given \(S = 25\),
\(1 = \sigma\sqrt{SC}\)). For relatively small systems (\(S \leq 25\)),
positive \(\sigma^{2}_{\gamma}\) never decreased the expected
\(\Re(\lambda_{max})\). But as \(S\) increased, the curvilinear
relationship between \(\rho\) and \(\Re(\lambda_{max})\) resulted in
\(\mathbf{M}\) with lower expected \(\Re(\lambda_{max})\) for low
magnitudes of \(\rho\). In turn, as \(S\) increased, and systems became
more complex, positive \(\sigma^{2}_{\gamma}\) increased the proportion
of numerical simulations that were observed to be stable (see below).

\textbf{Simulation of random \(\mathbf{M}\) across \(\mathbf{S}\)}. To
investigate the effect of \(\sigma^{2}_{\gamma}\) on stability across
systems of varying complexity, I simulated random \(\mathbf{M}\)
matrices at \(\sigma = 0.4\) and \(C = 1\) across
\(S = \{2, 3, ..., 49, 50\}\). One million \(\mathbf{M}\) were simulated
for each \(S\), and the stability of \(\mathbf{M}\) was assessed given
\(\gamma = 1\) (\(\sigma^{2}_{\gamma} = 0\)) versus
\(\gamma \sim \mathcal{U}(0, 2)\) (\(\sigma^{2}_{\gamma} = 1/3\)). For
all \(S > 10\), I found that the number of stable random systems was
higher given \(\sigma^{2}_{\gamma} > 0\) than when
\(\sigma^{2}_{\gamma} = 0\) (Fig. 3; see Supplementary Information for
full table of results), and that the difference between the
probabilities of observing a stable system increased with an increase in
\(S\). In other words, the potential for \(\sigma^{2}_{\gamma}\) to
affect stability increased with system complexity and was most relevant
for systems on the cusp of being too complex to be realistically stable.
For the highest values of \(S\), nearly all systems that were stable
given varying \(\gamma\) would not have been stable given
\(\gamma = 1\).

I also simulated 100000 \(\mathbf{M}\) for three types of random
networks that are typically interpreted as differing interspecific
ecological interaction
types\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2011}{25}}.
These interactions types are competitive, mutualist, or predator-prey,
as modelled by off-diagonal elements of \(\mathbf{M}\) that are
constrained to be negative, positive, or paired such that if
\(A_{ij} > 0\) then \(A_{ji} < 0\),
respectively\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(but are otherwise identical to the purely random \(\mathbf{M}\)). As
\(S\) increased, a higher number of stable systems were observed given
\(\sigma^{2}_{\gamma}>0\) versus \(\sigma^{2}_{\gamma}=0\) for
competitor and predator-prey, but not mutualist, systems. A higher
number of stable systems were observed whenever \(S > 12\) and
\(S > 40\) for competitive and predator-prey systems, respectively (note
that \(\rho < 0\) by definition for predator-prey systems, making
stability more likely overall). The stability of mutualist systems was
never affected by \(\sigma^{2}_{\gamma}\).

@@

\textbf{Simulation of structured \(\mathbf{M}\) across \(\mathbf{S}\)}.
To investigate how \(\sigma^{2}_{\gamma}\) affects the stability of
commonly observed network structures, I simulated 1 million
\(\textbf{M}\) for
small-world\textsuperscript{\protect\hyperlink{ref-Watts1998}{15}},
scale-free\textsuperscript{\protect\hyperlink{ref-Albert2002}{16}}, and
cascade food
web\textsuperscript{\protect\hyperlink{ref-Solow1998}{17},\protect\hyperlink{ref-Williams2000}{18}}
networks. In all of these networks, rules determining the presence or
absence of an interaction between components \(i\) and \(j\) constrained
the overall structure of the network. In small-world networks,
interactions between components are constrained so that the expected
degree of separation between any two components increases in proportion
to \(\log(S)\)\textsuperscript{\protect\hyperlink{ref-Watts1998}{15}}.
In scale-free networks, the number of components with which a focal
component interacts follows a power law; typically a few components have
many interactions while most components have few
interactions\textsuperscript{\protect\hyperlink{ref-Albert2002}{16}}. In
cascade food webs, species are ranked and interactions are constrained
such that a species \(i\) can only feed on \(j\) if the rank of
\(i > j\).

Network structure did not strongly modulate the effect that
\(\sigma^{2}_{\gamma}\) had on stability. For comparable magnitudes of
complexity, structured networks still had a higher number of stable
\(\mathbf{M}\) for \(\sigma^{2}_{\gamma}>0\) compared to
\(\sigma^{2}_{\gamma}=0\). For random networks, positive
\(\sigma^{2}_{\gamma}\) increased stability given \(S > 10\)
(\(\sigma = 0.4\) and \(C = 1\)), and therefore complexity
\(\sigma \sqrt{SC} \gtrapprox 1.26\). This threshold of complexity was
comparable for small-world networks, and slightly lower for scale-free
networks (note that algorithms for generating small-world and scale-free
networks necessarily led to varying \(C\), see methods). Varying
\(\gamma\) increased stability in cascade food webs for \(S > 27\), and
therefore at a relatively low complexity magnitudes compared to random
predator-prey networks. Overall, network structure does not greatly
change the effect that \(\sigma^{2}_{\gamma}\) has on increasing the
upper bound of complexity within which stability might reasonably be
observed.

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that values of all components are
positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{26},\protect\hyperlink{ref-Song2018}{27}}.
This is of particular interest for ecological communities because
population density (\(N\)) cannot take negative values, meaning that
ecological systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{26}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices \(\mathbf{M}\) tested for stability, and
additionally for specific types of ecological
communities\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(e.g., competitive, mutualist, predator-prey; see Supplementary
Information). I emphasise that \(\gamma\) is not interpreted as
population density in this analysis, but instead as a fundamental
property of species life history such as expected generation time.
Feasibility was unaffected by \(\sigma^{2}_{\gamma}\) and instead
occurred with a fixed probability of \(1/2^{S}\), consistent with a
recent proof by Serván et
al.\textsuperscript{\protect\hyperlink{ref-Servan2018}{28}} (see
Supplementary Information). Hence, for pure interacting species
networks, variation in component response rate (i.e., species generation
time) does not affect stability at biologically realistic species
densities.

@@

\textbf{Figure 1: Eigenvalue distributions of random complex systems.}
Each panel shows the real (x-axis) and imaginary (y-axis) parts of
\(S =\) 400 eigenvalues from random \(S \times S\) matrices.
(\(\textbf{a}\)) A system represented by a matrix \(\mathbf{A}\), in
which all elements are sampled from a normal distribution with
\(\mu = 0\) and \(\sigma_{A} = 1/\sqrt{S}\). Points are uniformly
distributed within a circle centred at the origin with a radius of
\(\sigma_{A} \sqrt{S} = 1\). (\(\textbf{b}\)) The same system after
including variation in the response rates of \(S\) components,
represented by the diagonal matrix \(\gamma\), such that
\(\mathbf{M} = \gamma\mathbf{A}\). Elements of \(\gamma\) are randomly
sampled from a uniform distribution from \(\min = 0\) to \(\max = 2\).
Eigenvalues of \(\mathbf{M}\) are then distributed non-uniformly within
a circle centred at the origin with a radius of
\(\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx\) 1.12.
(\(\textbf{c}\)) A different random system \(\mathbf{A}\) constructed
from the same parameters as in \(\textbf{a}\), except with diagonal
values of \(-1\). (\(\textbf{d}\)) The same system \(\textbf{c}\) after
including variation in component response rates, sampled from
\(\mathcal{U}(0, 2)\) as in \(\textbf{b}\).

>>>>

\textbf{Numerical simulations of random systems with and without
\(\mathbf{\sigma^{2}_{\gamma}}\)}. I used numerical simulations and
eigenanalysis to test how variation in \(\gamma\) affects stability in
random matrices with known properties, comparing the stability of
\(\textbf{A}\) versus \(\mathbf{M} = \gamma\mathbf{A}\). Values of
\(\gamma\) were sampled from a uniform distribution where
\(\gamma_{i} \sim \mathcal{U}(0, 2)\) and \(\sigma^{2}_{\gamma} = 1/3\)
(see Supplementary Information for other \(\gamma\) distributions, which
gave similar results). Diagonal elements of simulated \(\mathbf{A}\)
were standardised to the mean value of sampled \(\gamma\) so that
\(E[d]\) of \(\textbf{M}\) and \(\textbf{M}\) were guaranteed to be
identical. Here I focus on the effect of \(\gamma\) across values of
\(\rho\), and for increasing system sizes (\(S\)) in random and
structured networks. By increasing \(S\), the objective is to determine
the effect of \(\gamma\) as system complexity increases toward the
boundary at which stability is realistic for a finite system.

@@

Ten thousand numerical simulations of \(\mathbf{M}\) revealed that
\(\sigma^{2}_{\gamma}\) results in a nonlinear relationship between
\(\rho\) and \(\Re(\lambda_{max})\), which can sometimes increase the
stability of the system. Figure 2 shows a comparison of
\(\Re(\lambda_{max})\) across \(\rho\) values for \(\mathbf{A}\)
(\(\sigma^{2}_{\gamma} = 0\)) versus \(\mathbf{M}\)
(\(\sigma^{2}_{\gamma} = 1/3\)) given \(S = 25\), \(C = 1\), and
\(\sigma = 0.4\). For \(-0.4 \leq \rho \leq 0.7\) (shaded region of Fig.
2), expected \(\Re(\lambda_{max})\) was lower in \(\mathbf{M}\) than
\(\mathbf{A}\). For \(\rho \geq -0.1\), the lower bound of the range of
\(\Re(\lambda_{max})\) values also decreased given
\(\sigma^{2}_{\gamma}\), resulting in negative \(\Re(\lambda_{max})\) in
\(\mathbf{M}\) for \(\rho = -0.1\) and \(\rho = 0\). Hence, across a
wide range of system correlations, variation in the response rate of
system components had a stabilising effect.

The stabilising effect of \(\sigma^{2}_{\gamma}\) across \(\rho\)
increased with increasing \(S\). Figure 3 shows 10000 numerical
simulations of \(\mathbf{M}\) across increasing \(S\) given \(C = 1\)
and \(\sigma = 0.2\) (\(\sigma\) has been lowered here to better
illustrate the effect of \(S\); note that now given \(S = 25\),
\(1 = \sigma\sqrt{SC}\)). For relatively small systems (\(S \leq 25\)),
\(\sigma^{2}_{\gamma}\) never decreased the expected
\(\Re(\lambda_{max})\). But as \(S\) increased, the curvilinear
relationship between \(\rho\) and \(\Re(\lambda_{max})\) decreased
expected \(\Re(\lambda_{max})\) for \(\mathbf{M}\) given low magnitudes
of \(\rho\). In turn, as \(S\) increased, and systems became more
complex, \(\sigma^{2}_{\gamma}\) increased the proportion of numerical
simulations that were observed to be stable (see below).

\textbf{Simulation of random \(\mathbf{M}\) across \(\mathbf{S}\)}. To
investigate the effect of \(\sigma^{2}_{\gamma}\) on stability across
systems of varying complexity, I simulated random
\(\mathbf{M = \gamma A}\) matrices at \(\sigma = 0.4\) and \(C = 1\)
across \(S = \{2, 3, ..., 49, 50\}\). One million \(\mathbf{M}\) were
simulated for each \(S\), and the stability of \(\mathbf{A}\) vesus
\(\mathbf{M}\) was assessed given \(\gamma_{i} \sim \mathcal{U}(0, 2)\)
(\(\sigma^{2}_{\gamma} = 1/3\)). For all \(S > 10\), I found that the
number of stable random systems was higher in \(\mathbf{M}\) than
\(\mathbf{A}\) (Fig. 3; see Supplementary Information for full table of
results), and that the difference between the probabilities of observing
a stable system increased with an increase in \(S\). In other words, the
potential for \(\sigma^{2}_{\gamma}\) to affect stability increased with
system complexity and was most relevant for systems on the cusp of being
too complex to be realistically stable. For the highest values of \(S\),
nearly all systems that were stable given varying \(\gamma\) would not
have been stable given \(\gamma = 1\).

I also simulated 100000 \(\mathbf{M = \gamma A}\) for three types of
random networks that are typically interpreted as differing
interspecific ecological interaction
types\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2011}{25}}.
These interactions types are competitive, mutualist, or predator-prey,
as modelled by off-diagonal elements that are constrained to be
negative, positive, or paired such that if \(A_{ij} > 0\) then
\(A_{ji} < 0\),
respectively\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(but are otherwise identical to the purely random \(\mathbf{A}\)). As
\(S\) increased, a higher number of stable \(\mathbf{M}\) relative to
\(\mathbf{A}\) were observed for competitor and predator-prey, but not
mutualist, systems. A higher number of stable systems were observed
whenever \(S > 12\) and \(S > 40\) for competitive and predator-prey
systems, respectively (note that \(\rho < 0\) for predator-prey systems,
making stability more likely overall). The stability of mutualist
systems was never affected by \(\sigma^{2}_{\gamma}\).

@@

\textbf{Simulation of structured \(\mathbf{M}\) across \(\mathbf{S}\)}.
To investigate how \(\sigma^{2}_{\gamma}\) affects the stability of
commonly observed network structures, I simulated 1 million
\(\mathbf{M = \gamma A}\) for
small-world\textsuperscript{\protect\hyperlink{ref-Watts1998}{15}},
scale-free\textsuperscript{\protect\hyperlink{ref-Albert2002}{16}}, and
cascade food
web\textsuperscript{\protect\hyperlink{ref-Solow1998}{17},\protect\hyperlink{ref-Williams2000}{18}}
networks. In all of these networks, rules determining the presence or
absence of an interaction between components \(i\) and \(j\) constrained
the overall structure of the network. In small-world networks,
interactions between components are constrained so that the expected
degree of separation between any two components increases in proportion
to \(\log(S)\)\textsuperscript{\protect\hyperlink{ref-Watts1998}{15}}.
In scale-free networks, the number of components with which a focal
component interacts follows a power law; typically a few components have
many interactions while most components have few
interactions\textsuperscript{\protect\hyperlink{ref-Albert2002}{16}}. In
cascade food webs, species are ranked and interactions are constrained
such that a species \(i\) can only feed on \(j\) if the rank of
\(i > j\).

Network structure did not strongly modulate the effect that
\(\sigma^{2}_{\gamma}\) had on stability. For comparable magnitudes of
complexity, structured networks still had a higher number of stable
\(\mathbf{M}\) than \(\mathbf{A}\). For random networks,
\(\sigma^{2}_{\gamma}\) increased stability given \(S > 10\)
(\(\sigma = 0.4\) and \(C = 1\)), and therefore complexity
\(\sigma \sqrt{SC} \gtrapprox 1.26\). This threshold of complexity,
above which more \(\mathbf{M}\) than \(\mathbf{A}\) were stable, was
comparable for small-world networks, and slightly lower for scale-free
networks (note that algorithms for generating small-world and scale-free
networks necessarily led to varying \(C\); see methods). Varying
\(\gamma\) increased stability in cascade food webs for \(S > 27\), and
therefore at a relatively low complexity magnitudes compared to random
predator-prey networks (\(S > 40\)). Overall, network structure did not
greatly change the effect that \(\sigma^{2}_{\gamma}\) had on increasing
the upper bound of complexity within which stability might reasonably be
observed.

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that values of all components are
positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{26},\protect\hyperlink{ref-Song2018}{27}}.
This is of particular interest for ecological communities because
population density (\(N\)) cannot take negative values, meaning that
ecological systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{26}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices tested for stability. I emphasise that
\(\gamma\) is not interpreted as population density in this analysis,
but instead as a fundamental property of species life history such as
expected generation time. Feasibility was unaffected by
\(\sigma^{2}_{\gamma}\) and instead occurred with a fixed probability of
\(1/2^{S}\), consistent with a recent proof by Serván et
al.\textsuperscript{\protect\hyperlink{ref-Servan2018}{28}} (see
Supplementary Information). Hence, for pure interacting species
networks, variation in component response rate (i.e., species generation
time) does not affect stability at biologically realistic species
densities.

@@

\textbf{Figure 1: Eigenvalue distributions of random complex systems.}
Each panel shows the real (x-axis) and imaginary (y-axis) parts of
\(S =\) 400 eigenvalues from random \(S \times S\) matrices.
(\(\textbf{a}\)) A system represented by a matrix \(\mathbf{A}\), in
which all elements are sampled from a normal distribution with
\(\mu = 0\) and \(\sigma_{A} = 1/\sqrt{S}\). Points are uniformly
distributed within a circle centred at the origin with a radius of
\(\sigma_{A} \sqrt{S} = 1\). (\(\textbf{b}\)) The same system after
including variation in the response rates of \(S\) components,
represented by the diagonal matrix \(\gamma\), such that
\(\mathbf{M} = \gamma\mathbf{A}\). Elements of \(\gamma\) are randomly
sampled from a uniform distribution from \(\min = 0\) to \(\max = 2\).
Eigenvalues of \(\mathbf{M}\) are then distributed non-uniformly within
a circle centred at the origin with a radius of
\(\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx\) 1.18.
(\(\textbf{c}\)) A different random system \(\mathbf{A}\) constructed
from the same parameters as in \(\textbf{a}\), except with diagonal
values of \(-1\). (\(\textbf{d}\)) The same system \(\textbf{c}\) after
including variation in component response rates, sampled from
\(\mathcal{U}(0, 2)\) as in \(\textbf{b}\).
